Visualizing the Geography of Genetic Variants

### 2016-06-30 paper.tex
adding figures
added more text, nearly draft finished figures and started fig captions

One of the key characteristics of any genetic variant, beyond its
potential phenotypic effects or its frequency, is its geographic
distribution. The geographic distribution of a genetic variant can shed
light on where the variant first arose, in what populations spread
within, and in turn can help one learn about historical patterns of
migration, genetic drift, and natural selection. Collectively the
geographic distribution of genetic variants can help to explain how
populations have been related through time (e.g.~levels of gene flow and
divergence). For these reasons, visual inspection of geographic maps for
genetic variants is common practice in genetic studies. Here we develop
an interactive visualization tool for illuminating the geographic
distribution of genetic variants. Through an efficient RESTful API and
dynamic front-end the Geography of Genetic Variants (GGV) browser
rapidly provides maps of allele frequencies in populations distributed
across the globe.

@@

Genetics researchers often face the following problem: the researcher
has identified one or more genetic variants of interest (e.g.~from a
genome-wide association, eQTL, or pharmacogenomic study) and would like
to know the geographic distribution of the variant. For example, the
researcher may hope to address: 1) implications for genomic medicine
(e.g.~is a risk allele geographically localized to a certain patient
population?) \textbf{???}; 2) design of follow-up studies (e.g.~what
population should be studied to observe variant carriers?) \textbf{???};
or 3) the biology of the variant in question (e.g.~does the variant
correlate with a known environmental factor in a manner suggestive of
some geographically localized selection pressure?) {[}1{]}. A simple
geographic map of the distribution of a genetic variant can be
incredibly insightful for these questions.

Contemporary population genetics researchers are also faced with the
challenge of large, high dimensional datasets. For example, is not
uncommon for a researcher in human genetics to have a dataset comprised
of thousands of individuals measured at hundreds of thousands or even
millions of single nucleotide vairants (SNVs). One common approach to
visualizing such high-dimensional data is to compress the SNV dimensions
down to a small number of latent factors, using a method such as
principal components analysis {[}2{]}, {[}3{]}, a model-based clustering
method such as STRUCTURE {[}4{]} or ADMIXTURE {[}5{]}. While these
approaches are extremely valuable, researchers can use them to often
without inspecting the underlying variant data in more detail. A natural
approach to gaining more insight to the overall structure of a
population genetic dataset is to visually inspect what geographic
patterns arise in allele frequency maps.

Unfortunately, generating geographic allele frequency maps is
time-consuming for the average researcher as it requires a combination
of data wrangling methods {[}6{]} and geographic plotting techniques
that are unfamiliar to most. Our aim here is to produce a tailored
system for rapidly constructing informative geographic maps of allele
frequency variation.

Our work is inspired by past tools such as the ALFRED database {[}7{]}
and the maps available on the HGDP Selection browser {[}8{]}. One of us
(J.N.) developed the scripts for the HGDP Selection Browser maps using
The Generic Mapping Tools (GMT) {[}9{]}, a powerful system of geographic
plotting scripts with a sbustantial learning curve. The plots from the
HGDP Selection Browser have proved useful, and have appeared in research
articles e.g \textbf{???}, books e.g. {[}10{]}, and have been ported and
made avialable on the UCSC Genome Browser (available under the XX track
of the browser).

Reference datasets for population genetic variation has greatly expanded
since the release of the HGDP Illumina 650Y dataset {[}11{]} that formed
the basis of the HGDP Selection Browser maps. The most notable advance
is the publication of the 1000 Genomes phase 3 data {[}12{]} though
additional datasets are coming online {[}13{]}, {[}14{]}.

In addition, novel approaches for data visualization have become more
widely available. In particular, web-based visualization tools, such as
Data Driven Documents (d3.js), offer powerful approaches to realize
these aims through useful methods for interactivity, advantages of
software development in modern web-browsers, a large open-source
development community, and ease of share-ability {[}15{]}.

@@

The Geography of Genetic Variants browser (GGV) uses the SVG and mapping
utilities of d3.js to generate interactive frequency maps, allowing for
quick and dynamic displays of the geographic distribution of a genetic
variant.

In order to allow for easy access to large commonly used public genomic
datasets, such as the 1000 Genomes project, Human Genome Diversity
project and POPRES project, we have developed an SQL database and
RESTful API for querying allele frequencies by chromosome and position
or reference SNP identifier {[}Figure 1{]}.

@@

\caption[Figure caption]{Figure caption}\label{fig:figure1}
\end{figure}
\end{CodeChunk}

Users can query single variants by chromosome and position identifiers,
by rsids {[}16{]}, or users can simply choose a random variant from
within a dataset. While many applications require inspection of the
distribution of a specific variant, from our experience, it can be very
helpful to quickly view the geographic distribution of several randomly
chosen variants to quickly gain a sense of structure in a dataset. For
instance, in data with deep population subdivision, it is obvious in the
consistent patterns of differentiation observed across markers. We also
expect this will be useful in teaching contexts -- as it provides a
highly visual way for learners to understand human genetic variation.

After a query, the GGV displays the frequency of a variant in a given
population as a pie chart where each slice represents minor and major
allele frequencies. Pie charts are displayed as points positioned at the
population's region of origin where the projection/map-view is chosen
based off of the geographic proximity of populations present in a given
dataset {[}Figure 2{]}.

@@

@@

An additional challenge is that allele frequencies between variants
often differ greatly, sometimes by orders of magnitude in a single
dataset. This has not been an pervasive problem until recently, as most
population genetic samples were based on genotype arrays biased towards
variants that are common in human populations (5-50\% in minor allele
frequency). With the advent of next generation sequencing technologies
and large samples of thousands of individuals, it is common for datasets
to contain rare variants {[}12{]}, {[}17{]}, {[}18{]}.

@@

@@

@@

{[}Joe - worth saying more about the API structure and giving example
calls and structure of JSON?{]}

@@

Support for this work was provided by the NIH-BD2K initiative (1U01
CA198933-0). The authors would also like to thank XX for supportive
conversations.

\section*{Acknowledgements}\label{acknowledgements-1}
\addcontentsline{toc}{section}{Acknowledgements}

@@

\hypertarget{refs}{}
\hypertarget{ref-novembre2009spatial}{}
1. Novembre J, Di Rienzo A. Spatial patterns of variation due to natural
selection in humans. Nature Reviews Genetics. Nature Publishing Group;
2009;10: 745--755.

\hypertarget{ref-price2006principal}{}
2. Price AL, Patterson NJ, Plenge RM, Weinblatt ME, Shadick NA, Reich D.
Principal components analysis corrects for stratification in genome-wide
association studies. Nature genetics. Nature Publishing Group; 2006;38:
904--909.

\hypertarget{ref-patterson2006population}{}
3. Patterson N, Price AL, Reich D. Population structure and
eigenanalysis. PLoS genet. Public Library of Science; 2006;2: e190.

\hypertarget{ref-pritchard2000inference}{}
4. Pritchard JK, Stephens M, Donnelly P. Inference of population
structure using multilocus genotype data. Genetics. Genetics Soc
America; 2000;155: 945--959.

\hypertarget{ref-alexander2009fast}{}
5. Alexander DH, Novembre J, Lange K. Fast model-based estimation of
ancestry in unrelated individuals. Genome research. Cold Spring Harbor
Lab; 2009;19: 1655--1664.

\hypertarget{ref-furchedata}{}
6. Furche T, Gottlob G, Libkin L, Orsi G, Paton NW. Data wrangling for
big data: Challenges and opportunities.

\hypertarget{ref-rajeevan2011alfred}{}
7. Rajeevan H, Soundararajan U, Kidd JR, Pakstis AJ, Kidd KK. ALFRED: An
allele frequency resource for research and teaching. Nucleic acids
research. Oxford Univ Press; 2011; gkr924.

\hypertarget{ref-pickrell2009signals}{}
8. Pickrell JK, Coop G, Novembre J, Kudaravalli S, Li JZ, Absher D, et
al. Signals of recent positive selection in a worldwide sample of human
populations. Genome research. Cold Spring Harbor Lab; 2009;19: 826--837.

\hypertarget{ref-wessel2013generic}{}
9. Wessel P, Smith WH, Scharroo R, Luis J, Wobbe F. Generic mapping
tools: Improved version released. Eos, Transactions American Geophysical
Union. Wiley Online Library; 2013;94: 409--410.

\hypertarget{ref-dudley2013exploring}{}
10. Dudley JT, Karczewski KJ. Exploring personal genomics. Oxford
University Press; 2013.

\hypertarget{ref-li2008worldwide}{}
11. Li JZ, Absher DM, Tang H, Southwick AM, Casto AM, Ramachandran S, et
al. Worldwide human relationships inferred from genome-wide patterns of
variation. science. American Association for the Advancement of Science;
2008;319: 1100--1104.

\hypertarget{ref-10002015global}{}
12. Consortium 1GP, others. A global reference for human genetic
variation. Nature. Nature Publishing Group; 2015;526: 68--74.

\hypertarget{ref-meyer2012high}{}
13. Meyer M, Kircher M, Gansauge M-T, Li H, Racimo F, Mallick S, et al.
A high-coverage genome sequence from an archaic denisovan individual.
Science. American Association for the Advancement of Science; 2012;338:
222--226.

\hypertarget{ref-lazaridis2014ancient}{}
14. Lazaridis I, Patterson N, Mittnik A, Renaud G, Mallick S, Kirsanow
K, et al. Ancient human genomes suggest three ancestral populations for
present-day europeans. Nature. Nature Publishing Group; 2014;513:
409--413.

\hypertarget{ref-bostock2011d3}{}
15. Bostock M, Ogievetsky V, Heer J. D\(^3\) data-driven documents.
Visualization and Computer Graphics, IEEE Transactions on. IEEE;
2011;17: 2301--2309.

\hypertarget{ref-sherry2001dbsnp}{}
16. Sherry ST, Ward M-H, Kholodov M, Baker J, Phan L, Smigielski EM, et
al. DbSNP: The nCBI database of genetic variation. Nucleic acids
research. Oxford Univ Press; 2001;29: 308--311.

\hypertarget{ref-nelson2012abundance}{}
17. Nelson MR, Wegmann D, Ehm MG, Kessner D, Jean PS, Verzilli C, et al.
An abundance of rare functional variants in 202 drug target genes
sequenced in 14,002 people. Science. American Association for the
Advancement of Science; 2012;337: 100--104.

\hypertarget{ref-tennessen2012evolution}{}
18. Tennessen JA, Bigham AW, O'Connor TD, Fu W, Kenny EE, Gravel S, et
al. Evolution and functional impact of rare coding variation from deep
sequencing of human exomes. Science. American Association for the
Advancement of Science; 2012;337: 64--69.

>>>>

One of the key characteristics of any genetic variant, beyond its
potential phenotypic effects or its frequency, is its geographic
distribution. The geographic distribution of a genetic variant can shed
light on where the variant first arose, in what populations it spread
within, and in turn can help one learn about historical patterns of
migration, genetic drift, and natural selection. Collectively the
geographic distribution of genetic variants can help to explain how
populations have been related through time (e.g.~levels of gene flow and
divergence). For these reasons, visual inspection of geographic maps for
genetic variants is common practice in genetic studies. Here we develop
an interactive visualization tool for illuminating the geographic
distribution of genetic variants. Through an efficient REST API and
dynamic front-end the Geography of Genetic Variants (GGV) browser
rapidly provides maps of allele frequencies in populations distributed
across the globe.

@@

Genetics researchers often face the following problem: the researcher
has identified one or many genetic variants of interest (e.g.~from a
genome-wide association, eQTL, or pharmacogenomic study) and would like
to know the geographic distribution of the variant. For example, the
researcher may hope to address: 1) implications for genomic medicine
(e.g.~is a risk allele geographically localized to a certain patient
population?); 2) design of follow-up studies (e.g.~what population
should be studied to observe variant carriers?); or 3) the biology of
the variant in question (e.g.~does the variant correlate with a known
environmental factor in a manner suggestive of some geographically
localized selection pressure?) {[}1{]}, {[}2{]}, {[}3{]}. A simple
geographic map of the distribution of a genetic variant can be
incredibly insightful for these questions.

The geographic distribution of rare genetic variants (minor allele
frequency \(\leq .05\)) can be particularly informative in determining
an individual's ancestry. Rare variants tend to be localized in small
geographic regions because they have arisen recently in human history
and have had little time to spread across populations through migration
{[}4{]}, {[}5{]}. Thus by surveying the collection of rare variants that
an individual carries provides information about the population origins
of that individual's recent ancestors. One of us (J.N.) recently took
advantage of the geographic clustering of an individual's rare variants
in helping to reaffirm the historically highly contested Native American
ancestry of the Kennewick Man, an ancient individual from Washington
State \textbf{???}.

Contemporary population genetics researchers are also faced with the
challenge of large, high dimensional datasets. For example, is not
uncommon for a researcher in human genetics to have a dataset comprised
of thousands of individuals measured at hundreds of thousands or even
millions of single nucleotide vairants (SNVs). One common approach to
visualizing such high-dimensional data is to compress the SNV dimensions
down to a small number of latent factors, using a method such as
principal components analysis {[}6{]}, {[}7{]}, a model-based clustering
method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While these
approaches are extremely valuable, researchers can use them to often
without inspecting the underlying variant data in more detail. A natural
approach to gaining more insight to the overall structure of a
population genetic dataset is to visually inspect what geographic
patterns arise in allele frequency maps.

Unfortunately, generating geographic allele frequency maps is
time-consuming for the average researcher as it requires a combination
of data wrangling methods {[}10{]} and geographic plotting techniques
that are unfamiliar to most. Our aim here is to produce a tailored
system for rapidly constructing informative geographic maps of allele
frequency variation.

Our work is inspired by past tools such as the ALFRED database {[}11{]}
and the maps available on the HGDP Selection browser {[}12{]}. One of us
(J.N.) developed the scripts for the HGDP Selection Browser maps using
The Generic Mapping Tools (GMT) {[}13{]}, a powerful system of
geographic plotting scripts with a sbustantial learning curve. The plots
from the HGDP Selection Browser have proved useful, and have appeared in
research articles e.g {[}12{]}, {[}14{]}, books e.g. {[}15{]}, and have
been ported and made avialable on the UCSC Genome Browser (available
under the HGDP Allele Freq track of the browser).

Reference datasets for population genetic variation has greatly expanded
since the release of the HGDP Illumina 650Y dataset {[}16{]} that formed
the basis of the HGDP Selection Browser maps. The most notable advance
is the publication of the 1000 Genomes phase 3 data {[}17{]} though
additional datasets are coming online {[}18{]}, {[}19{]}.

In addition, novel approaches for data visualization have become more
widely available. In particular, web-based visualization tools, such as
Data Driven Documents (D3.js), offer powerful approaches to realize
these aims through useful methods for interactivity, advantages of
software development in modern web-browsers, a large open-source
development community, and ease of share-ability {[}20{]}.

@@

The Geography of Genetic Variants browser (GGV) uses the SVG and mapping
utilities of D3.js to generate interactive frequency maps, allowing for
quick and dynamic displays of the geographic distribution of a genetic
variant.

@@

\caption[Schematic diagram of the Geography of Genetic Variants (GGV) web application]{Schematic diagram of the Geography of Genetic Variants (GGV) web application. We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genom Diversity Project (HGDP) and Population Reference Sample (POPRES). These allele freqeuncies are stored in a SQL database along side with geographic coordinates associated with each population. The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}\label{fig:figure1}
\end{figure}
\end{CodeChunk}

In order to allow for easy access to large commonly used public genomic
datasets, such as the 1000 Genomes project, Human Genome Diversity
project and POPRES project, we have developed an SQL database and REST
API for querying allele frequencies by chromosome and position or
reference SNP identifier {[}Figure 1{]}.

Users can query single variants by chromosome and position identifiers,
by rsid {[}21{]}, or users can simply choose a random variant from
within a dataset. While many applications require inspection of the
distribution of a specific variant, from our experience, it can be very
helpful to quickly view the geographic distribution of several randomly
chosen variants to quickly gain a sense of structure in a dataset. For
instance, in data with deep population subdivision, it is obvious in the
consistent patterns of differentiation observed across markers. We also
expect this will be useful in teaching contexts -- as it provides a
highly visual way for learners to understand human genetic variation.

{\centering \includegraphics{paper_files/figure-latex/figure2-1}

\caption[Example map from the Geography of Genetic Variants web application]{Example map from the Geography of Genetic Variants web application. Each pie chart represents a population with the blue slice of the pie displaying the freqeuncy of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:figure2}
\end{figure}
\end{CodeChunk}

After a query, the GGV displays the frequency of a variant in a given
population as a pie chart where each slice represents minor and major
allele frequencies. Pie charts are displayed as points positioned at the
population's region of origin where the projection / map-view is chosen
based off of the geographic proximity of populations present in a given
dataset {[}Figure 2{]}.

@@

{\centering \includegraphics{paper_files/figure-latex/figure3-1}

\caption[Figure caption]{Figure caption}\label{fig:figure3}
\end{figure}
\end{CodeChunk}

@@

An additional challenge is that allele frequencies between variants
often differ greatly, sometimes by orders of magnitude in a single
dataset. This has not been an pervasive problem until recently, as most
population genetic samples were based on genotype arrays, biased towards
variants that are common in human populations (5-50\% in minor allele
frequency). With the advent of next generation sequencing technologies
and large samples of thousands of individuals, it is common for datasets
to contain rare variants {[}17{]}, {[}4{]}, {[}22{]}.

@@

{\centering \includegraphics{paper_files/figure-latex/figure4-1}

\caption[Figure caption]{Figure caption}\label{fig:figure4}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{paper_files/figure-latex/figure5-1}

\caption[Figure caption]{Figure caption}\label{fig:figure5}
\end{figure}
\end{CodeChunk}

@@

To interface population minor allele frequency data stored in the SQL
database with the front-end D3.js visualization we use a REST API
impelemented in the python library Flask-RESTful \textbf{???}. GET
requests are submitted to the database via HTTP, returning json
formatted allele frequency data and the meta-data associated with each
population and genetic variant e.g.~latitude, longitude, population
label, sample size, and frequency scale. Genetic variants can be queried
by chromosome position, rsid, or randomly. Example HTTP requests and
json response can be seen below:

\subsection*{Query API by rsid}\label{query-api-by-rsid}
\addcontentsline{toc}{subsection}{Query API by rsid}

\begin{verbatim}
http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&rsID=rs1834640
[
  {
    "alleles": ["A", "G"], 
    "pos": ["-15.310139", "13.443182"], 
    "pop": "GWD", 
    "nobs": "226", 
    "xobs": "17", 
    "freqscale": 1, 
    "freq": [0.0752212389381, 0.9247787610619], 
    "chrom_pos": "15:48392165", 
    "rawfreq": 0.0752212389381
  },
  ...
]
\end{verbatim}

\subsection*{Query API by chromosome
position}\label{query-api-by-chromosome-position}
\addcontentsline{toc}{subsection}{Query API by chromosome position}

\begin{verbatim}
http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&chr=14&pos=37690093
[
  {
    "alleles": ["G", "A"], 
    "pos": ["-15.310139", "13.443182"], 
    "pop": "GWD", 
    "nobs": "226", 
    "xobs": "0", 
    "freqscale": 0.01, 
    "freq": [0.0, 1.0], 
    "chrom_pos": "14:37690093", 
    "rawfreq": 0.0
  }
  ...
]
\end{verbatim}

\subsection*{Randomly query API}\label{randomly-query-api}
\addcontentsline{toc}{subsection}{Randomly query API}

\begin{verbatim}
http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&random_snp=True
[
  {
    "alleles": ["T", "C"], 
    "pos": ["-15.310139", "13.443182"], 
    "pop": "GWD", 
    "nobs": "226", 
    "xobs": "0", 
    "freqscale": 0.01, 
    "freq": [0.0, 1.0], 
    "chrom_pos": "5:42452893", 
    "rawfreq": 0.0
  }, 
  ...
]
\end{verbatim}

@@

Support for this work was provided by the NIH-BD2K initiative (1U01
CA198933-0). The authors would also like to thank members of the
Novembre Lab for supportive conversations.

@@

\hypertarget{refs}{}
\hypertarget{ref-rosenberg2010genome}{}
1. Rosenberg NA, Huang L, Jewett EM, Szpiech ZA, Jankovic I, Boehnke M.
Genome-wide association studies in diverse populations. Nature Reviews
Genetics. Nature Publishing Group; 2010;11: 356--366.

\hypertarget{ref-novembre2009spatial}{}
2. Novembre J, Di Rienzo A. Spatial patterns of variation due to natural
selection in humans. Nature Reviews Genetics. Nature Publishing Group;
2009;10: 745--755.

\hypertarget{ref-coop2010using}{}
3. Coop G, Witonsky D, Di Rienzo A, Pritchard JK. Using environmental
correlations to identify loci underlying local adaptation. Genetics.
Genetics Soc America; 2010;185: 1411--1423.

\hypertarget{ref-nelson2012abundance}{}
4. Nelson MR, Wegmann D, Ehm MG, Kessner D, Jean PS, Verzilli C, et al.
An abundance of rare functional variants in 202 drug target genes
sequenced in 14,002 people. Science. American Association for the
Advancement of Science; 2012;337: 100--104.

\hypertarget{ref-mathieson2014demography}{}
5. Mathieson I, McVean G. Demography and the age of rare variants. PLoS
Genet. Public Library of Science; 2014;10: e1004528.

\hypertarget{ref-price2006principal}{}
6. Price AL, Patterson NJ, Plenge RM, Weinblatt ME, Shadick NA, Reich D.
Principal components analysis corrects for stratification in genome-wide
association studies. Nature genetics. Nature Publishing Group; 2006;38:
904--909.

\hypertarget{ref-patterson2006population}{}
7. Patterson N, Price AL, Reich D. Population structure and
eigenanalysis. PLoS genet. Public Library of Science; 2006;2: e190.

\hypertarget{ref-pritchard2000inference}{}
8. Pritchard JK, Stephens M, Donnelly P. Inference of population
structure using multilocus genotype data. Genetics. Genetics Soc
America; 2000;155: 945--959.

\hypertarget{ref-alexander2009fast}{}
9. Alexander DH, Novembre J, Lange K. Fast model-based estimation of
ancestry in unrelated individuals. Genome research. Cold Spring Harbor
Lab; 2009;19: 1655--1664.

\hypertarget{ref-furchedata}{}
10. Furche T, Gottlob G, Libkin L, Orsi G, Paton NW. Data wrangling for
big data: Challenges and opportunities.

\hypertarget{ref-rajeevan2011alfred}{}
11. Rajeevan H, Soundararajan U, Kidd JR, Pakstis AJ, Kidd KK. ALFRED:
An allele frequency resource for research and teaching. Nucleic acids
research. Oxford Univ Press; 2011; gkr924.

\hypertarget{ref-pickrell2009signals}{}
12. Pickrell JK, Coop G, Novembre J, Kudaravalli S, Li JZ, Absher D, et
al. Signals of recent positive selection in a worldwide sample of human
populations. Genome research. Cold Spring Harbor Lab; 2009;19: 826--837.

\hypertarget{ref-wessel2013generic}{}
13. Wessel P, Smith WH, Scharroo R, Luis J, Wobbe F. Generic mapping
tools: Improved version released. Eos, Transactions American Geophysical
Union. Wiley Online Library; 2013;94: 409--410.

\hypertarget{ref-coop2009role}{}
14. Coop G, Pickrell JK, Novembre J, Kudaravalli S, Li J, Absher D, et
al. The role of geography in human adaptation. PLoS Genet. Public
Library of Science; 2009;5: e1000500.

\hypertarget{ref-dudley2013exploring}{}
15. Dudley JT, Karczewski KJ. Exploring personal genomics. Oxford
University Press; 2013.

\hypertarget{ref-li2008worldwide}{}
16. Li JZ, Absher DM, Tang H, Southwick AM, Casto AM, Ramachandran S, et
al. Worldwide human relationships inferred from genome-wide patterns of
variation. science. American Association for the Advancement of Science;
2008;319: 1100--1104.

\hypertarget{ref-10002015global}{}
17. Consortium 1GP, others. A global reference for human genetic
variation. Nature. Nature Publishing Group; 2015;526: 68--74.

\hypertarget{ref-meyer2012high}{}
18. Meyer M, Kircher M, Gansauge M-T, Li H, Racimo F, Mallick S, et al.
A high-coverage genome sequence from an archaic denisovan individual.
Science. American Association for the Advancement of Science; 2012;338:
222--226.

\hypertarget{ref-lazaridis2014ancient}{}
19. Lazaridis I, Patterson N, Mittnik A, Renaud G, Mallick S, Kirsanow
K, et al. Ancient human genomes suggest three ancestral populations for
present-day europeans. Nature. Nature Publishing Group; 2014;513:
409--413.

\hypertarget{ref-bostock2011d3}{}
20. Bostock M, Ogievetsky V, Heer J. D\(^3\) data-driven documents.
Visualization and Computer Graphics, IEEE Transactions on. IEEE;
2011;17: 2301--2309.

\hypertarget{ref-sherry2001dbsnp}{}
21. Sherry ST, Ward M-H, Kholodov M, Baker J, Phan L, Smigielski EM, et
al. DbSNP: The nCBI database of genetic variation. Nucleic acids
research. Oxford Univ Press; 2001;29: 308--311.

\hypertarget{ref-tennessen2012evolution}{}
22. Tennessen JA, Bigham AW, O'Connor TD, Fu W, Kenny EE, Gravel S, et
al. Evolution and functional impact of rare coding variation from deep
sequencing of human exomes. Science. American Association for the
Advancement of Science; 2012;337: 64--69.
